---
title: "154HW1_Jiyoon_Jeong_Lab2"
author: "Jiyoon Clover Jeong"
date: "9/10/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1

```{r}

A <- matrix(c(1,4,2,0,-3,1),2,3)
B <- matrix(c(2,-1,3,2,4,0),2,3)
C <- matrix(c(0,4,1,-1,0,-2),2,3)

A + B
(A + C) + B
A -(C + B)
-(A + B)
(A - B) + C

```


## Problem 2

```{r}

X <- matrix(c(2, 1, 0, 9, 4, 2, 3, 8, 3, 5, 2, 4, 7, 3, 4, 5, 8, 7, 7, 2, 9, 8, 7, 1), 6, 4, byrow = T)

rownames(X) <- c(letters[1:6])
colnames(X) <- c("Y","X1","X2","X3")
X


```


### (a)
```{r}

X[1,1] + X[2,1] + X[3,1] + X[4,1] + X[5,1] + X[6,1]

X[,1] %*% matrix(1, 6, 1)


```


### (b)

```{r}


(X[1,2] + X[2,2] + X[3,2] + X[4,2] + X[5,2] + X[6,2]) /6

(X[,2] %*% matrix(1, 6, 1)) /6


```


### (c)

```{r}

t(X[,1, drop =F]) %*% X[,3, drop = F]

```


### (d)


```{r}

(t(X[,4, drop = F]) %*% X[,4, drop = F] ) - (t(X[,4, drop = F]) %*% matrix(1,6,1))^(2)/6

```


### (e)

```{r}

meany <- (X[,1] %*% matrix(1, 6, 1)) /6
meanx1 <- (X[,2] %*% matrix(1, 6, 1)) /6
meanx2 <- (X[,3] %*% matrix(1, 6, 1)) /6
meanx3 <- (X[,4] %*% matrix(1, 6, 1)) /6

cbind((X[,1,drop =F] - meany[1]), (X[,2,drop =F] - meanx1[1]),
           (X[,3,drop =F] - meanx2[1]), (X[,4,drop =F] - meanx3[1]))
X

# check with : apply(X,2, FUN = function(x) {x - mean(x)})
```

### (f)

```{r}
p <- ncol(X) #number of variables
n <- nrow(X) #number of objects
 
#create means for each column
meanmat <- matrix(data=1, nrow=n) %*% cbind(meany[1],meanx1[1],meanx2[1],meanx3[1]) 
 
#creates a difference matrix
D <- X - meanmat
 
#creates the covariance matrix
Cov <- (n-1)^(-1) * (t(D) %*% D)
Cov

# check with cov(X)
cov(X)

```

## problem 3

### (a)

```{r}

a <- abs(0.5)
b <- abs(4)

a * b * cos(45*pi/180)



```

### (b)

```{r}

a <- abs(4)
b <- abs(1)

a * b * cos(90*pi/180)

```

### (c)

```{r}

a <- abs(1)
b <- abs(1)

a * b * cos(120*pi/180)

```


## problem 4

```{r}

u = c(1, 3, 5)
v = c(2, 4, 6)

# v onto u
proj <- function(v,u){
  
  c <- (t(u) %*% v) / (t(u) %*% u)
  return(c*u)
  
}
proj(u,v)

```


## problem 5
### (a)

```{r}
x = c(1, 2, 3)
y = c(3, 0, 2)
z = c(3, 1, 1)

norm <- function(x){
  return(sum(x^2)^(1/2))
}

u1 <- x
e1 <- u1/ norm(u1)
cat("u1 : ", u1, "\n")
cat("e1 : ", e1, "\n")

u2 <- y - proj(y,u1)
e2 <- u2 / norm(u2)
cat("u2 : ", u2, "\n")
cat("e2 : ", e2, "\n")

u3 <- z - proj(z,u1) - proj(z,u2)
e3 <- u3 / norm(u3)
cat("u3 : ", u3, "\n")
cat("e3 : ", e3, "\n")



```


### (b)

```{r}

x = c(2, 1)
y = c(1, 2)
z = c(1, 1)

u1 <- x
e1 <- u1/ norm(u1)
cat("u1 : ", u1, "\n")
cat("e1 : ", e1, "\n")

u2 <- y - proj(y,u1)
e2 <- u2 / norm(u2)
cat("u2 : ", u2, "\n")
cat("e2 : ", e2, "\n")

u3 <- z - proj(z,u1) - proj(z,u2)
u3 <- round(u3,10)
e3 <- u3 / norm(u3)
cat("u3 : ", u3, "\n")
cat("Since u3 is (0, 0), we can omit u3 and conclude that e1 and e2 forms orthonormal basis of R^2. As we know, basis of R^2 have list which length is 2 so it is impossible to have three list of basis for R^2.")


```
## problem 6

```{r}

lp_norm <- function(x, p=1){
  if(p =="max"){
    return(max(abs(x)))
  }
  else{
    lp <- sum((abs(x)^p))^(1/p)
  }
  return(lp)
}



zero <- rep(0, 10)
p = 1
lp_norm(zero,p)

ones <- rep(1, 5)
p = 3
lp_norm(ones,p)

u <- rep(0.4472136, 5)
p = 2
lp_norm(u,p)

u <- c(-40:0)
p = 100
lp_norm(u,p)

u <- c(1:1000)
p = "max"
lp_norm(u, p)
```


## problem 7

```{r}
u1 <- 1/sqrt(11) * c(3, 1, 1)
u2 <- 1/sqrt(6) * c(-1, 2, 1)
u3 <- 1/sqrt(66) * c(-1,-4, 7)

#norms of u1, u2, and u3 are zero.
lp_norm(u1,2)
lp_norm(u2,2)
lp_norm(u3,2)

#<u1,u2> = <u1,u3> = <u2,u3> = 0
t(u1) %*% u2
t(u1) %*% u3

t(u2) %*% u3


```

From the calculation above, u1, u2, and u3 are orthonormal vectors

## problem 8

### (a)

```{r}

X <- as.matrix(USArrests)
class(X)

```


### (b)
```{r}

n= nrow(X)
p = ncol(X)

cat("# rows: ", n, "\n")
cat("# columns: ", p, "\n")


```


### (c)
```{r}

D <- diag(1/n,n)
sum(diag(D))

```



### (d) column means
```{r}

one <- rep(1,n)
g <- t(X) %*% D %*% one
g

```


### (e)  mean-centered matrix
```{r}
one <- matrix(1,n,1)
Xc <- X - one %*% t(g)
colMeans(Xc)

```


### (f) the (population) variance-covariance matrix
```{r}

V <- t(X) %*% D %*% X - g %*% t(g)
V

```


### (g) D_s be a p×p diagonal matrix with elements on the diagonal equal to 1/Sj , where Sj is the standard deviation for the j-th variable
```{r}

D_s <-diag(c(1/sd(X[,1]),1/sd(X[,2]),1/sd(X[,3]), 1/sd(X[,4]))) 

D_s <-diag(diag(1/sqrt(V)))
D_s
diag(D_s)

```


### (h) matrix of standardized data
```{r}


Z <- Xc %*% D_s

colMeans(Z)
apply(Z, 2, sd)

```


### (i) the (population) correlation matrix
```{r}

R <- D_s %*% V %*% D_s
R

```


### (j) R can also be obtained as R = t(Z) D Z
```{r}

R <- t(Z) %*% D %*% Z
R

cor(X)

```


