---
title: "154Lab8"
author: "Jiyoon Clover Jeong"
date: "10/23/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(ISLR)
library(ggplot2)
library(FactoMineR)

names(Default)
dim(Default)
summary(Default)
summary(subset(Default, default == 'Yes'))
summary(subset(Default, default == 'No'))


ggplot(data = Default, aes(x = balance, y = income, color = default)) + geom_point(alpha = 0.4) + labs(title = "Scatterplot between Balance and Income")


ggplot(data = Default, aes(x = balance, fill = default, color = default)) +  geom_density(alpha = 0.4)


ggplot(data = Default, aes(x = income, fill = default, color = default)) + geom_density(alpha = 0.4)

```

# OLS Regression

```{r}
default_numeric <- rep(0, nrow(Default))
default_numeric[Default$default == 'Yes'] <- 1
Default$default_num <- default_numeric
ols_reg <- lm(default_num ~ balance, data = Default)

summary(ols_reg)



# Q
ggplot(data = Default, aes(x = balance, y = default_num)) + geom_smooth(method = "lm", se = F) + geom_point(aes(col = default)) 



#aes(x = Default$balance, y = Default$default_num)



```

# Logistic Regression

```{r}


logreg_default <- glm(default ~ balance, family = binomial, data = Default)
summary(logreg_default)$coefficients
logreg_default



newdata = data.frame(balance = seq(100,2000,100))
predict(logreg_default, newdata, type="response")

logreg_default <- glm(default ~ student, family = binomial, data = Default)
summary(logreg_default)


logreg_default <- glm(default ~ balance + student + income, family = binomial, data = Default)
summary(logreg_default)
print("income coefficient is not significant")



```


# The Stock Market Smarket Data

```{r}
names(Smarket)
dim(Smarket)
summary(Smarket)


cor <- cor(Smarket[,-9])
cor
PCA(Smarket[,-9])

# plot(x = Smarket$Year, y = Smarket$Volume)
# lines(lowess(Year,Volume), col="blue")


ggplot(data = Smarket, aes(x = Year, y = Volume)) + geom_point() + geom_smooth(method = loess)



```


# Logistic Regression

```{r}
formula <- paste0("Lag", 1:5, collapse = " + ") 
formula <- paste("Direction ~", formula, "+ Volume")
fit <- glm(formula, family = binomial, data = Smarket)

summary(fit)



# Q : if new data ommited, fitted value will be used 
head(predict(fit, type = "response"))

```

Lag1 seems to be the most significant. The coefficient of Lag1 is -0.073074 and has nagative sign.



# Estimation of Parameters

• Let y be the column vector of response Y
• Let X be the n × (p + 1) input (design) matrix
• Let p be the n-vector of fitted probabilities with the i-th element $p(xi, \beta^{old})$
• Let W be an n×n diagonal matrix of weights with i-th element $p(xi, \beta^{old})(1-p(xi, \beta^{old}))$


# Newton-Raphson algorithm

```{r}

newdirect <- rep(0, nrow(Smarket))
newdirect[Smarket$Direction == "Up"] <- 1
Smarket$newdirect <- newdirect
n = nrow(Smarket)



y <- as.matrix(Smarket[,10, drop = F])
X <- as.matrix(cbind(1, Smarket[, c(2:7), drop = F]))
p <- c(0)
W <- diag(0,n)
b_old <- matrix(0, ncol(X), 1)
b_new <- matrix(0, ncol(X), 1)
diff <- 10^10

while(diff >10^(-7)){
   b_old <- b_new
  for(j in 1:n){
    p[j] <- exp( X[j,, drop = F] %*% b_old )   /
      (1 + exp( X[j, , drop = F] %*% b_old ) )

    W[j,j] = p[j] * (1- p[j])

  }

  z <- X %*% b_old + solve(W) %*% (y - p)
  b_new <- solve(t(X) %*% W %*% X ) %*% t(X) %*% W %*% z
  
  diff <- sqrt(sum((b_new - b_old)^2))

}

b_new
 
```

# Simplified Algorithm

```{r}

b_old <- matrix(0, ncol(X), 1)
y <- as.matrix(Smarket[,10, drop = F])
X <- as.matrix(cbind(1, Smarket[, c(2:7), drop = F]))
p <- c(0)
diff <- 10^10


while(diff >10^(-7)){
  b_old <- b_new
  for(j in 1:n){
    p[j] <- exp( X[j, , drop = F] %*% b_old )   /
      (1 + exp( X[j, , drop = F] %*% b_old ) )
    
  }
  X_hat <- sweep(X, MARGIN = 1, p * (1-p), FUN = '*') 
  
  b_new <- b_old + solve(t(X) %*% X_hat) %*% t(X) %*% (y-p)
  
  diff <- sqrt(sum((b_new - b_old)^2))
  
}

b_new




```



```{r}


```



```{r}


```



```{r}


```



```{r}


```


