---
title: "154HW5"
author: "Jiyoon Clover Jeong"
date: "10/26/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nnet)
library(ggplot2)
library(caret)

```


```{r}
wine <- read.table("/Users/cloverjiyoon/2017Fall/Stat 154/data/wine.data.txt", sep = ",", header =  T)
wine$class <- as.factor(wine$class)

names(wine)


```

# 1) Sum-of-Squares Dispersion Functions (10 pts)


```{r}

tss <- function(x) {
  
  sum((x - mean(x))^2)
}

cat("TSS : ", tss(iris$Sepal.Length), "\n")



bss <- function(x, y){
  
  
  y <- as.factor(as.character(y))
  
  if(length(x) != length(y))
    stop("x and y have different lengths")
  
  
  sum <- 0
  x_bar <-  mean(x)      #tss(x)
  splited <- split(x, y)
  
  X_k <- sapply(splited, mean)
  
  for(i in 1: length(splited)){
    sum <- sum + length(splited[[i]]) * (X_k[i] - x_bar)^2
  }
  
  return(sum)
  
}

cat("BSS : ", bss(iris$Sepal.Length, iris$Species), "\n")



wss <- function(x,y){
  
  y <- as.factor(as.character(y))

  if(length(x) != length(y))
    stop("x and y have different lengths")

  sum <- 0
  x_bar <-  mean(x)      #tss(x)
  splited <- split(x, y)
  X_k <- sapply(splited, mean)

  for(i in 1: length(splited)){
    for(j in 1: length(splited[[i]])){
       sum <- sum +  (splited[[i]][j] - X_k[i])^2
    }
   
  }

  return(sum)

}


cat("WSS : ", wss(iris$Sepal.Length, iris$Species), "\n")



```

# 2) Sum-of-Squares Ratio Functions (10 pts)

```{r}

cor_ratio <- function(x, y){
  BSS <- bss(x,y)
  TSS <- tss(x)
  
  cor <- BSS / TSS
  return (as.numeric(cor))
}

cor_ratio(iris$Sepal.Length, iris$Species)


F_ratio <- function(x,y){
  
  y <- as.factor(as.character(y))
  BSS <- bss(x,y)
  WSS <- wss(x,y)
  
  n <- length(x)
  k <- nlevels(y)
    
  Fratio <- (BSS / (k - 1)) / (WSS / (n - k))
  return(as.numeric(Fratio))
}



F_ratio(iris$Sepal.Length, iris$Species)


```


# 3) Discriminant Power of Predictors (30 pts)

```{r}

# The first approach consists of running simple logistic regressions
# Q : For  number 3, consider only wine classes 1 and 2 (ignore class 3)

predictors <- names(wine[1:130,-1])
formula <- c("class ~")
AIC <- data.frame(name = predictors, AIC = rep(0,13))

for(i in 1: length(predictors)){
  
  formula <- paste("class ~", predictors[i])
  #fit <- multinom(formula, data = wine[1:130,])
  fit <- glm(formula, data = wine[1:130,] , family = binomial)
  AIC[i,2] <- fit$aic
  
  
}

AIC

AIC <- AIC[ order(AIC[,2]), ]

AIC

ggplot(AIC, aes( x = reorder(name, AIC), y =  AIC)) + geom_col() + theme(axis.text.x = element_text(angle = 60, hjust = 1))




# The second approach consists of computing correlation ratios

corratio <- data.frame(name = predictors, cor_ratio = rep(0,13))

for(i in 1:length(predictors)){
  corratio[i,2] <- cor_ratio(wine[1:130,i+1], wine[1:130,1])
}

corratio

corratio <- corratio[ order( corratio[,2]), ]

corratio


# Q : automatic ordering?? 
#ggplot(corratio, aes(x = name, y = cor_ratio)) + geom_col()

ggplot(corratio, aes(x = reorder(name, cor_ratio), y = cor_ratio)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 60, hjust = 1))


# Q : automatic ordering?? 
#ggplot(corratio, aes(x = name, y = cor_ratio)) + geom_bar(stat = "identity")




# The third approach consists of computing F-ratios

Fratio <- data.frame(name = predictors, Fratio = rep(0,13))

for(i in 1:length(predictors)){
  Fratio[i,2] <- F_ratio(wine[1:130,i+1], wine[1:130,1])
}

Fratio

Fratio <- Fratio[ order( Fratio[,2]), ]

Fratio

ggplot(Fratio, aes(x = reorder(name, Fratio), y = Fratio)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 60, hjust = 1))




```


# 4) Variance functions


```{r}

total_variance <- function(X){
  X <- as.matrix(X)
  X <- scale(X, scale = F)
  n = dim(X)[1]
  V <- (1/ (n-1)) * t(X) %*% X
  return(V)
  
}

# test total_variance()
total_variance(iris[ ,1:4])

# compare with var()
var(iris[ ,1:4])



# PPT 25- p64

between_variance <- function(X, y){
  
  X <- scale(X, scale = F)
  n <- length(y)
  y <- as.data.frame( y)
  Y <- model.matrix(~ y -1, data =  y)
  BSS <- t(X) %*% Y %*% solve(t(Y) %*% Y)  %*% t(Y)  %*% X
  B <- BSS / (n - 1)
  
  return(B)
  
}


# test between_variance()
between_variance(iris[ ,1:4], iris$Species)


# PPT 25- p64

within_variance <- function(X, y){
  
  # X <- as.data.frame(scale(X, scale =F))
  # #X <- scale(X, scale = F)
  # n <- dim(X)[1]
  # p <- dim(X)[2]
  # XX <- data.frame(X,y)
  # X_k <-  split(X, y)
  # GSS_k <- list()
  # W_k <- list()
  # W <- matrix(0, p, p)
  # for(i in 1:nlevels(y)){
  #   X_k[[i]] <- as.matrix(X_k[[i]])
  #   GSS_k[[i]] <- t(X_k[[i]]) %*% X_k[[i]]
  #   n_k <- dim(X_k[[i]])[1]
  #   W_k <- (1 / (n_k -1)) *  GSS_k[[i]]
  #   W <- W + ( (n_k - 1) / (n - 1)) * W_k
  # }
  
  X <- scale(X, scale = F)
  n <- length(y)
  y <- as.data.frame(y)
  Y <- model.matrix(~ y -1, data =  y)
  WSS <- t(X) %*% (diag(n) - Y %*% solve(t(Y) %*% Y) %*% t(Y)) %*% X
  W <- WSS / (n - 1)

  
  return(W)
}


# test within_variance()
within_variance(iris[ ,1:4], iris$Species)



# Confirm that V = B +W

# confirm V = B + W
Viris <- total_variance(iris[ ,1:4])
Viris


# B + W
Biris <- between_variance(iris[ ,1:4], iris$Species)
Wiris <- within_variance(iris[ ,1:4], iris$Species)
Biris + Wiris



```



# Challenge

## find the eigenvectors uk

```{r}

# find the eigenvectors uk

X <- wine[,2:14]
y <- as.factor(wine[,1])

W <- within_variance(X, y)    # X should be dataframe

K <- nlevels(y)
J <- dim(X)[2]
n <- dim(X)[1]

splited <- split(X, y)

C <- matrix(0, J, K)

for(j in 1: J){
  for(k in 1: K){
    
    n_k <- dim(splited[[k]])[1]
    n_k
    Xbar_jk <- mean(splited[[k]][, j])
    Xbar_jk
    Xbar_j <- mean(X[, j])
    Xbar_j
    C[j,k] <- sqrt(n_k / (n-1)) *  (Xbar_jk - Xbar_j)
    C[j,k]
    
  }
}

C

# check

B <- between_variance(X,y)


head(B,2)
head(C %*% t(C), 2)

dim(B)

dim(C %*% t(C))


eigen_w <- eigen( t(C) %*% solve(W) %*% C )$vectors

eigen_values <- eigen( t(C) %*% solve(W) %*% C )$values

eigen_values

eigen_u <- solve(W) %*% C %*% eigen_w

eigen_u




```


## Obtain the linear combinations zk and make a scatterplot of the wines

```{r}

X <- as.matrix(X)

Z <- X %*% eigen_u
head(Z)
Z_splited <- split(as.data.frame(Z), y)


# for(i in 1:nlevels(y)){
#   z_k <- c(Z_splited[[1]][, i], Z_splited[[2]][ ,i],  Z_splited[[3]][ ,i])
#   Z_k <- data.frame(z_k, y)
#   print(ggplot(Z_k, aes(x = z_k, y = y, color = y)) + geom_point() +
#           labs(title = paste0("linear combinations Z_", i), y = "class", color = "class"))       
#   
# }




  z_k <- rbind(Z_splited[[1]][, 1:2], Z_splited[[2]][ ,1:2], 
           Z_splited[[3]][ ,1:2])
 
  Z_k <- data.frame(z_k, y)
  print(ggplot(Z_k, aes(x = V1, y = V2, color = y)) + geom_point())   
  
  
  


```

## scatterplot of the wines but this time using the first two principal components on the standardized predictors.

```{r}

X_scaled <- scale(X)

score <-princomp(X_scaled, cor = TRUE)$score[,1:2]

score[,1:2]


n1 <- dim(Z_splited[[1]])[1]
n2 <- dim(Z_splited[[2]])[1]


score <- cbind(score, y = y)
score


ggplot(as.data.frame(score), aes(x = Comp.1, y = Comp.2, color = as.factor(y))) + geom_point()



# for(i in 1:2){
#   z_k <- c(Z_splited[[1]][, i], Z_splited[[2]][ ,i],  Z_splited[[3]][ ,i])
#   Z_k <- data.frame(z_k, y)
#   print(ggplot(Z_k, aes(x = z_k, y = y, color = y)) + geom_point() +
#           labs(title = paste0("linear combinations Z_", i), y = "class", color = "class"))       
#   
# }



```


## Calculate the correlations between zk and the predictors

```{r}

# Q : Z with standardzied predictors or not??
# Q : interpret score?

cor(Z[,-3], X)



```


we can think 


## Create a matrix of size n × K, with the squared Mahalanobis distances

```{r}

mahal <- matrix(0, n, K)

X_splited <- split(as.data.frame(X), y)

g_k <- lapply(X_splited, colMeans)


W_inverse <- solve(W)

for(i in 1:n){
  for(j in 1:K){
    mahal[i,j] <- (X[i, , drop = F] - g_k[[j]])  %*% W_inverse %*%
      t(X[i, , drop = F] - g_k[[j]])
  }
}

head(mahal)


```


### 1. assign each observation to the class Gk for which the Mahalanobis distance d^2(xi, gk) is the smallest

```{r}


assigned <- apply(mahal, 1, which.min)

assigned

```


### 2. create a confussion matrix comparing the actual class versus the predicted class
```{r}


#data.frame(assigned, y)


table(assigned, y)


confusionMatrix(assigned, y)

```

