---
title: "154Lab4"
author: "Jiyoon Clover Jeong"
date: "9/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

```{r}


reg <-lm(mpg ~ disp, data = mtcars)
mtcars_c <- as.data.frame(scale(mtcars, center = T, scale = F))
reg2 <-lm(mpg ~ disp, data = mtcars_c)
mtcars_c <- as.data.frame(scale(mtcars, center = T, scale = T))
reg3 <-lm(mpg ~ disp, data = mtcars_c)

lm(mpg ~ disp -1, data = mtcars)

# mtcars has a column am for automatic transmission: 0 = automatic, 1 = manual. Find out how to use the argument subset of lm() to regress mpg on disp, with just those cars having automatic transmission.

lm(mpg ~ disp, data = mtcars, subset = am == 0)
am_0 <- mtcars %>% filter(am == 0)
lm(mpg ~ disp, am_0)

```



```{r}
reg_sum <- summary(reg)
reg_sum
class(reg_sum)
names(reg_sum)

```




```{r}

plot(mtcars$disp, mtcars$mpg)
abline(reg, lwd = 2)

plot.new()
plot(mtcars$disp, mtcars$mpg)
abline(reg, col = "Red", lty = 4, lwd = 2) #lty is a line type and lwd is line width

# scatterplot with regression line


# Here’s how to get a nicer plot using low-level plotting functions:
plot.new()
plot.window(xlim = c(50, 500), ylim = c(10, 40))
points(mtcars$disp, mtcars$mpg, pch = 20, cex = 1)
abline(reg, col = "tomato", lwd = 2) # regression line
axis(side = 1, pos = 10, at = seq(50, 500, 50))
axis(side = 2, las = 1, pos = 50, at = seq(10, 40, 5))



ggplot(data = mtcars, aes(x = disp, y = mpg)) + geom_point() + geom_text(label = rownames(mtcars), size = 3, alpha = 0.5) +  xlim(50,500) + ylim(10, 35) + geom_smooth(method = "lm", se = FALSE) 

```



```{r}
#residuals vs fitted model
plot(reg, which = 1)

# normal probability plot
plot(reg, which = 2)

reg_anova <- anova(reg)
reg_anova


```



# beta coefficients
XtXi <- solve(t(X) %*% X)
XtXi %*% t(X) %*% y

solve(crossprod(X, X), crossprod(X, y))

=======================================


Q is an n × n orthogonal matrix
 R is a p × p upper triangular matrix
QR <- qr(X)

To extract the matrices of the decomposition, you have to use the companion functions qr.Q() and qr.R()
Q <- qr.Q(QR)
R <- qr.R(QR)
f <- t(Q %*% y)    # y?

Solving Rb = f we use the method of backsubstitution:
backsolve(R, f)



# Recover coefficients
## Normal Equation

```{r}


X <- as.matrix(cbind(1,mtcars[,3]))
y <- as.matrix(mtcars[,"mpg"])

lm(mpg ~ disp, data = mtcars)
beta <- solve(t(X) %*% X) %*% t(X) %*% y

```


## QR Decomposition

```{r}

QR <- qr(X)

Q <- qr.Q(QR)
R <- qr.R(QR)
f <- t(Q) %*% y
backsolve(R, f)

```


## Solve function 

```{r}

solve(t(X) %*% X, t(X) %*% y )

solve(R, f)
solve(R, t(Q) %*% y)

```

